---
title: "Reproducible Research: Peer Assessment 1"
author: "Vern"
date: "8/15/2015"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

### Loading

The dataset is here: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)  

The dataset is an csv-formatted file, with data collected over the period Oct 01 thru Nov 31 2012, at 5-min intervals throughout each day.  

There are 17568 observations with three variables:  

1. **steps**: an integer vector with missing values stored as NAs

2. **date**: a character vector of dates formated as, YYYY-mm-dd

3. **interval**: a character vector of 5-min intervals over a 24 hour period  
  

We can download, unzip, and read in the file as per the R code below:   

```{r, message=FALSE, eval=TRUE}
# load libraries
library(utils)      # for download.file()

## assuming the name of the needed file in the dataset is "activity.csv"

# create temp file, and download dataset to temp file (method='curl' for 'https' url)
temp <- tempfile()
URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(URL, temp, method = 'curl')

# read in data; remove temp file
activity_data <- read.csv(unz(temp, "activity.csv"))
unlink(temp)

```

[`tempfile`](http://www.rdocumentation.org/packages/base/functions/tempfile),
[`download.file`](http://www.rdocumentation.org/packages/utils/functions/download.file), [`read.csv`](http://www.rdocumentation.org/packages/utils/functions/read.table), 
[`unlink`](http://www.rdocumentation.org/packages/base/functions/unlink)  
__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------

### Data Preprocessing

Two variables will be added to the dataset:  

1. **date_times**, a datetime vector of the dates and times combined  

2. **interval_seq**: a times vector of the 5-minute intervals for a 24-hour period

```{r, echo=TRUE, message=FALSE }
# load libraries
library(dplyr)      ## manipulate the data
library(lubridate)  ## for ymd_hm function
library(chron)      ## for times class 

# convert dataframe to tbl_df (dplyr)
activity_data <- tbl_df(activity_data)

# add variable 'date_times' (POSIXct class) to dataset
activity_data <- activity_data %>%
     mutate(date_times = seq(from = ymd_hm("2012.10.1 0:0", tz = "UTC"), to = ymd_hm("2012.11.30 23:55"), by = 300))

# create char vector of sequence of 5-minute intervals over a 24-hour period
interval_seq <- format(seq(c(ISOdatetime(2012,10,1,0,0,0)), by = 300, length.out = 288), "%H:%M:%S")

# add 'interval_seq' to dataset; convert from char to times vector
activity_data$interval_seq <-interval_seq
activity_data <- activity_data %>%
     mutate(interval_seq = chron(times. = interval_seq))

```

[`dplyr`](http://www.rdocumentation.org/packages/dplyr),
[`lubridate`](http://www.rdocumentation.org/packages/lubridate/functions/lubridate), [`chron`](http://www.rdocumentation.org/packages/chron/functions/chron)  

__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------

## What is mean total number of steps taken per day?

We calculate the summary statistics by grouping by date.   

We use [`ggvis`](http://www.rdocumentation.org/packages/ggvis/functions/ggvis) to plot the histogram:  

```{r, echo=TRUE, message=FALSE}
#load libraries
library(ggvis)

# daily totals of steps taken per day
daily_totals <- activity_data %>%
     group_by(d = cut.POSIXt(date_times, breaks = "days")) %>%
     summarise(daily_sum = sum(steps, na.rm = TRUE))

# daily mean and daily median
daily_mean <- round(mean(daily_totals$daily_sum, na.rm = TRUE))
daily_median <- median(daily_totals$daily_sum, na.rm = TRUE)

# Histogram of total number of steps taken daily
daily_totals %>%
     ggvis(~daily_sum) %>%
     layer_histograms(width=2000, fill := "blue") %>%
     add_axis("x", title = "Number of Steps") %>%
     add_axis("y", title = "Frequency")

```


The mean of the total number of steps taken per day: **``r daily_mean``**

The median of the total number of steps taken per day: **``r daily_median``**  

__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------

## What is the average daily activity pattern?

This time we summarise by the intervals to calculate the daily activity patterns:

```{r, echo=TRUE}

# time series plot of the 5-minute interval vs the average number of 
# steps taken, averaged across all days (y-axis)
interval_means <- activity_data %>%
     group_by(interval) %>%
     summarise(interval_mean_steps = mean(steps, na.rm = TRUE))

interval_means %>%
     ggvis(~interval, ~interval_mean_steps) %>%
     layer_lines(stroke := "blue") %>%
     add_axis("x", title = "24-hour Time (5-Minute Intervals)", format = "d") %>%
     add_axis("y", title = "Mean Steps per Interval")

# 5-minute interval with maximum number of steps on average
max_steps_interval <- activity_data$interval_seq[which.max(interval_means$interval_mean_steps)]

```


The 5-minute interval with the maximum number of steps on average: **``r as.character(max_steps_interval)``**     

__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------

## Imputing missing values

Given that there are missing values (recorded as NAs) in the dataset, this may introduce bias into some calculations or summaries of the data.  

* The total number of missing values (NAs) in the dataset: **``r sum(is.na(activity_data))``**

* All missing values are recorded only in the **steps** variable.

```{r, echo=TRUE, result=FALSE}

var_NAs <- sapply(activity_data, function(y) sum(length(which(is.na(y)))))

print(var_NAs)

```


We will use the mean value for the relevant 5-minute interval to fill in the corresponding missing values.  

```{r, echo=TRUE}
 
# we replace NAs by interval mean value for the particular interval time slot
# add the interval_mean_steps to activity_data: round the value, and convert to integer
activity_data$interval_mean_steps <- as.integer(round(interval_means$interval_mean_steps))

# add variable, updated_steps
# if steps == NA, replace with relevant interval_mean_steps for the interval: 
activity_data <- activity_data %>% 
     mutate(imp_steps = ifelse(is.na(steps), interval_mean_steps, steps))

```


We create a new dataset equal to the original dataset with the NAs replaced by interval means.

```{r, echo=TRUE}

imp_activity_data <- activity_data %>% select(imp_steps, date_times, interval)

```


With the updated dataset, the histogram now looks like below:  

```{r, echo=TRUE}

# update calculations with imputed values
imp_daily_totals <- imp_activity_data %>%
     group_by(d = cut.POSIXt(date_times, breaks = "days")) %>%
     summarise(imp_daily_sum = sum(imp_steps, na.rm = TRUE))

# updated daily mean and daily median
imp_daily_mean <- round(mean(imp_daily_totals$imp_daily_sum, na.rm = TRUE))
imp_daily_median <- median(imp_daily_totals$imp_daily_sum, na.rm = TRUE)

# updated histogram of total number of steps taken daily
imp_daily_totals %>%
     ggvis(~imp_daily_sum) %>%
     layer_histograms(width=2000, fill := "blue") %>%
     add_axis("x", title = "Number of Steps (with imputed values)") %>%
     add_axis("y", title = "Frequency(with imputed values)")

```

The mean of the total number of steps taken per day: **``r as.integer(round(imp_daily_mean))``**

The median of the total number of steps taken per day: **``r imp_daily_median``**


We may note that:  

* the distribution for the updated histogram looks a little more like a normal distribution  

* the difference between the mean and median for the updated dataset is minimal 

An interesting question here would be: how do we validate the choice of the interval mean as a fill for missing values?

__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------

## Are there differences in activity patterns between weekdays and weekends?

It would seem that this is the case.  From the panel plot below, we can see a difference in the activity for the weekdays against the weekends.  On average, it looks like there is a spike in activity all week long around the 08:00 to 10:00 intervals.  However, on average, weekends appear to show more activity during the normal daytime hours.   

For processing the weekday vs weekend, we use the `is.weekend` function from the `chron` library   

For the panel plot, we use the [`lattice`](http://www.rdocumentation.org/packages/lattice/functions/Lattice) library  

We will add a new factor variable, **week_days** to the updated dataset with two levels: “weekday” and “weekend,” which will indicate whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
# load libraries
library(lattice)

# add new variables, 'week_days' and 'imp_interval_mean', to dataset
imp_activity_data <- imp_activity_data %>%
     mutate(week_days = ifelse(is.weekend(date_times), "weekend", "weekday"))

imp_activity_data$week_days <- factor(imp_activity_data$week_days)

imp_mean_steps_per_int <- imp_activity_data %>%
     group_by(interval, week_days) %>%
     summarise(imp_interval_mean = mean(imp_steps, na.rm = TRUE))

# generate panel plot
xyplot((imp_interval_mean ~ interval | week_days), 
       data = imp_mean_steps_per_int,
       type = "l",
       xlab = "Interval",
       ylab = "Number of steps",
       layout=c(1,2))

```

__________________________________________________________________________________________________________________

----------------------------------------------------------------------------------